Post-replication citation patterns in psychology: Four case studies

### 2021-02-06 manuscript/manuscript.Rmd
completed transfer of introduction from google doc to R markdown

bibliography      : ["references.bib"]

@@

```{r patterns}
tibble(
  "Progressive responses" = c(1, 2),
  "Regressive responses" = c(1, 2)
) %>%
  apa_table(caption = "Progressive or regressive responses to strongly contradictory replication results and their expected impact on citation patterns for original studies.")
```

>>>>

@@

# Introduction

It is often assumed that science is a self-correcting enterprise: the veracity of scientific knowledge should progressively improve as inaccurate claims are abandoned and accurate claims are reinforced (Vazire & Holcombe, 2020). Replication studies are considered to be a key driver of this process because they may indicate that prior results are exaggerated or erroneous (Ioannidis, 2012; Zwaan et al., 2018). Although interpreting the outcome of replication studies is not necessarily straightforward (Collins, 1985; Earp & Trafimow, 2015; Maxwell et al., 2015), one might expect a replication result that strongly contradicts[^footnote1] and outweighs the results of a prior (‘original’) study to impact how that study is cited in subsequent academic literature. For example, if a replication undermines belief in the credibility of an original finding, one might expect to see a change in the frequency and valence (favourability) of citations to the original study; specifically, a decrease in favourable citations accompanied by an increase in unfavourable citations. However, as discussed below, a variety of interesting patterns could emerge depending on how the research community responds to a replication result. The goal of the present study was to empirically explore and describe post-replication citation patterns in the context of four prominent multi-laboratory replication attempts published in the field of psychology that strongly contradicted and outweighed the findings of prior studies.

[^footnote1]: Determining whether a replication finding contradicts an original finding is a complex issue which we largely side-step here by only examining cases that are strongly contradictory - that is, according to several reasonable quantitative criteria (e.g., effect size magnitude, p-values, Bayes Factors, etc.) the results of the original study and replication study lead to opposing inferences (e.g., absence vs. presence of an effect). Also note that one can accept that results are strongly contradictory and remain agnostic about the explanation for the contradiction.

Table \@ref(tab:patterns) outlines several citation patterns that might follow a contradictory replication result, each reflecting different types of response by the research community. We have tentatively categorised these response patterns as being ‘progressive’ or ‘regressive’ depending on their expected impact on the accumulation of scientific knowledge[^footnote2]. The first set of patterns, belief correction/perpetuation, refer to what is often considered the primary functional role of (contradictory) replication studies - to change belief in the credibility of exaggerated or erroneous original findings (Ioannidis, 2012; Vazire & Holcombe, 2020; Zwaan et al., 2018). In the absence of a explicit defence of an original finding that convincingly explains a strongly contradictory result that outweighs an original finding (see ‘explicit defence’ below), a progressive response might involve a decrease in favourable citations and an increase in unfavourable citations, reflecting updated beliefs about the credibility of the original finding. Conversely, a regressive response might involve maintenance of (or even increase in) favourable citations, and relatively few unfavourable citations, suggesting a perpetuation of belief in the credibility of the original finding despite the contradictory replication result. Prior research has documented how favourable citations to observational epidemiology studies can persist despite the claims of those studies being strongly contradicted in subsequent randomized trials (Tatsioni et al., 2007). Similarly, it has been reported that even when articles are retracted, they can continue to receive favourable citations (Budd et al., 1998). Thus, there is evidence that belief in the credibility of original findings can perpetuate even when subsequent events cast doubt on their credibility; however, we are unaware of similar evidence in the context of replication studies that were explicitly designed to test the reliability of prior findings.

[^footnote2]: This terminology was inspired by, but does not directly mirror, terminology proposed by Imre Lakatos in his work on the rationality of the research community’s response when a scientific theory is contracted by empirical evidence (Lakatos, 1970).

The second set of patterns in Table \@ref(tab:patterns), citation balance/bias, generally refers to whether positive (supportive) evidence is preferentially cited relative to negative (non-supportive) evidence. This pattern has previously been observed in the context of research on inclusion body myositis; citation content analysis showed that the accumulating literature heavily cited the theory that beta amyloid is involved, ignoring multiple studies that contradicted this theory (Greenberg, 2009; also see Bastiaansen et al., 2015). In the present study, these patterns specifically refer to whether articles citing an original study also cite the subsequent contradictory replication study. A progressive response would be to cite both studies (citation balance), as this involves considering and reporting highly relevant evidence (even if the implications of the replication are disputed, see ‘explicit defence’ below). By contrast, citation bias could occur if articles citing an original study neglect to cite a relevant replication study. Regardless of whether this occurs through lack of awareness or deliberate omission, it can be considered a regressive response because highly relevant evidence is not being reported or considered.

The third set of patterns in Table \@ref(tab:patterns), explicit/absent defence, refers to whether researchers who continue to favourably cite the original finding despite the strongly contradictory replication result offer a concrete defence of the original study. As implied above, even when the result of a replication study strongly contradicts the results of an original study, this does not necessarily nullify the credibility of the original findings; the same criticisms that one might apply to an original study in order to infer that its findings are erroneous or exaggerated may also be applied to replication studies (Collins, 1985; Earp & Trafimow, 2015; Maxwell et al., 2015). Thus, if proponents of the original claim mount an explicit defence that counters the implications of the replication results, this might still be considered a progressive response (though obviously one could disagree with the arguments that are presented). By contrast, if favourable citations to the original study are not accompanied by explicit argumentation about the replication result (an absent defence) this might be considered a regressive response because the replication result is apparently discounted without providing any rationale.

```{r patterns}
tibble(
  "Progressive responses" = c("*Belief correction*
A decrease in the number of favourable citations may reflect a decline in belief in the credibility of the original finding (a belief correction pattern). This may be accompanied by a relative increase in the number of unfavourable citations (an active belief correction pattern) or relatively fewer/no unfavourable citations (a passive belief correction pattern).", "*Citation balance*
Articles that cite the original study may also cite the contradictory replication, reflecting that relevant evidence has been considered (a balanced citation pattern).", "*Explicit defence*
Articles that favourably cite the original study and unfavourably cite the contradictory replication may offer concrete counter-arguments that state why the credibility of the original finding has not been undermined (an explicit defence pattern)."),
  "Regressive responses" = c("*Belief perpetuation*
A maintenance of, or increase in, the number of favourable citations may reflect that a maintenance of or increase in belief in the credibility of the original finding (a belief perpetuation pattern). This may be accompanied by a relative increase in unfavourable citations (a challenged belief perpetuation pattern) or relatively fewer/no unfavourable citations (an unchallenged belief perpetuation pattern).", "*Citation bias*
Articles that cite the original study may neglect to also cite the contradictory replication, reflecting that relevant evidence has been neglected, either through lack of awareness or deliberate omission (a citation bias pattern).", "*Absent defence*
Articles that favourably cite the original study and unfavourably cite the contradictory replication may offer no concrete counter-arguments that state why the credibility of the original finding has not been undermined (an absent defence pattern).")
) %>%
  kable(booktabs=T,longtable=T,caption = "Progressive or regressive responses to strongly contradictory replication results and their expected impact on citation patterns for original studies.") %>%
  column_spec(1,'8.5cm') %>%
  column_spec(2,'8.5cm')
```

In the present study, we explored the post-replication citation patterns described above in the context of four case studies in the field of psychology where the findings of a replication study strongly contradicted and outweighed the findings of an original study (Table \@ref(tab:studyDetails)). In two of the cases, the replication studies were part of a single ‘Many Labs’ project (Klein et al., 2014) and addressed the ‘flag priming effect’ (T. J. Carter et al., 2011) and ‘money priming effect’ (Caruso et al., 2013) respectively. The other two cases involved ‘Registered Replication Reports’ (RRRs; Simons et al., 2014), examining influential demonstrations of the ‘facial feedback effect’ (cf. Strack et al., 1988; Wagenmakers et al., 2016) and the ‘ego-depletion effect’ (cf. Baumeister et al., 1998; Hagger et al., 2016; Sripada et al., 2014) respectively. For methodological reasons (see Hagger et al., 2016), the ego-depletion replication was aimed at a classic study in the field (Baumeister et al., 1998), but actually employed a modified computer-based version of the original paradigm (Sripada et al., 2014). Thus, for this particular case study we examined citation patterns to both of these original studies. We chose these particular case studies because they involved prominent pre-registered multi-laboratory replication attempts with sample sizes between 23 - 211 times larger than the original studies, thus providing highly visible and highly credible evidence that strongly contradicted and outweighed earlier findings[^footnote3]. This facilitates additional interpretative clarity about the citation patterns one might expect to observe.

[^footnote3]: We deliberately focused on a select group of case studies rather than other potentially larger samples in order to aid interpretative clarity; for example, the extent to which the results of the large scale Reproducibility Project in Psychology (RPP; Open Science Collaboration, 2015) actually contradicted the original studies has been contested (Etz & Vandekerckhove, 2016). Additionally, original studies were not actually cited in the RPP research report, which may have diluted awareness about relevant replication studies.

The present study was exploratory in nature and intended to provide descriptive observations rather than test hypotheses. The three sets of expected citation patterns outlined in Table \@ref(tab:patterns) were used to guide our study design and interpretation, but we do not claim that this is a comprehensive typology of the post-replication patterns that may occur. Such patterns may be more complex and idiosyncratic in other topic domains. In order to examine patterns of belief correction/perpetuation, we downloaded citation histories (a list of citing articles) for each original study and classified the valence of a set of pre-replication and post-replication citations as favourable, equivocal, or unfavourable. To examine patterns of citation balance/bias, we manually checked whether post-replication citations of the original study were accompanied by citations to the replication study. Finally, to examine patterns of explicit/absent defence, we extracted and categorised any counter-arguments offered in articles that cited the replication.
